Human Factors Influencing Visual Statistical Inference

### 2013-06-15 sociology_paper.tex
model discussion

\subsection{Learning Trend} Models \eqref{eqn:trend_response} and \eqref{eqn:trend_time} are fitted to the data from experiment 5, 6, and 7 separately. Table \ref{tbl:model_result_response} presents the parameter estimates and $p$-values of fixed effect estimates of Model \eqref{eqn:trend_response}. The larger $p$-values suggest that none of the levels of attempt ($\alpha_2$ through $\alpha_{10}$) appears to be significant at \%1 significance level. Moreover, some of the estimates are positive and some are negative and they show up seemingly in random order suggesting later attempts not necessarily improved. This indicates that there may be no learning effect on the probability of correct evaluations.

@@

The estimated variance ($\sigma^2_a$) of the random effect attempt is much small compared to other random effects except for subject variability ($\sigma^2_u$) for experiment 7. This suggest that most of the variability for attempt is accounted for by the fixed effect factors of attempt. For experiment 7 the lineup variability ($\sigma^2_l$) was the largest as per experimental design. Some of the lineups were very easy and some were really difficult in experiment 7. The overall average probability for picking out the data plot from a lineup ($\mu$) appears to be significant for both experiments 5 and 7. But for experiment 6 it does not seem to be significant suggesting that experiment 6 lineups were difficult compared to the other experimental lineups. This shows the designs of experiments such that there exists the mixture of difficult and easy lineups within the experiments as well as between the experiments.  For easy lineups there may be small chance to improve performances. But for harder lineups, the scope to improve performances is large.  Since for none of these experiments improvement in performances is observed, it is important in the sense that performances did not improve over attempts in both difficult and easy lineup situations.

@@

To visualize further how the performance in correct response improves over successive attempt, we fitted Model \eqref{eqn:trend_response} excluding the covariates related to attempt from the model and computed the residuals. The averages of these residuals for each of the attempts are plotted against attempt as shown in Figure \ref{fig:learning_trend_response}. Least square linear regression lines are drown through the points. Even though positive slopes over the attempts are observed, none of the trends is statistically significant.

Table \ref{tbl:model_result_time} presents the results of Model \eqref{eqn:trend_time}. The parameter $\alpha$ for fixed effect covariate attempt is highly significant in all the experiments. The negative estimates suggest that later attempt took less time for an evaluation. This suggest that even though observers did not improve the performance over attempt, they became efficient in responding faster. The parameter $\alpha_1$ for first attempt is also highly significant. The positive estimates of $\alpha_1$ indicates that first attempt made by an observer required much more times than other attempts. It is because for initial attempt the observer might have gone through instructions and became familiar with the experimental environment. Each page in the web site contains information about choice reason and the observers's confidence level. Also, the first page asks for observer Identification to be typed. The later pages of the web site was similar just the lineup was changed. Thus in the later attempts an observer does not need to spend any time for reading instructions. The model reflects that fact.

@@

\subsection{How People Pick the Data Plot} In the experiment setup we have a question for the turk observers to choose the reason for their selection of a particular data plot. It is shown by \cite{majumder:2013} that these choice reasons reveal the way people picks the data plot in the lineup. To investigate this further we added free text input option for their choice reason instead of some fixed reasons to select from. In this experiment people could write whatever they think their reasons for choice are. Figure \ref{fig:wordle} shows the words used to explain their reasons for choice. The most common words used to explain their choice are points and green which indicates two important feature of a plot. One is the indicator of plot aesthetics and the other is the color. Spread, steepest, line and apart are some other important words used frequently. Spread and apart are indicative of variability in the data. Steepest and line indicates some sort of systematic pattern in the data.

>>>>

\subsection{Learning Trend} Models \eqref{eqn:trend_response} and \eqref{eqn:trend_time} are fitted to the data from experiment 5, 6, and 7 separately.  As an alternative to Model \eqref{eqn:trend_response} we examined a reduced model with attempt as a continuous covariate. But that reduced model appears to be significantly different with a $p$-value of XXX. Thus we consider the bigger Model \eqref{eqn:trend_response} with attempt as a factor variable with 10 levels. Attempt 1 is considered to be base level while fitting the model.

Table \ref{tbl:model_result_response} presents the parameter estimates and $p$-values of fixed effect estimates of Model \eqref{eqn:trend_response}. The larger $p$-values suggest that none of the levels of attempt ($\alpha_2$ through $\alpha_{10}$) appears to be significant at \%1 significance level. Moreover, some of the estimates are positive and some are negative and they show up seemingly in random order suggesting later attempts not necessarily improved. This indicates that there may be no learning effect on the probability of correct evaluations.

@@

The variance of random slope for attempt ($\sigma^2_a$) is estimated to be very small compared to other random effects except for subject variability ($\sigma^2_u$) for experiment 7. This suggest that most of the variability for attempt is accounted for by the fixed effect factors of attempt. For experiment 7 the lineup variability ($\sigma^2_l$) was the largest as per experimental design. Some of the lineups were very easy and some were really difficult in experiment 7. The overall average probability for picking out the data plot from a lineup ($\mu$) appears to be significant for both experiments 5 and 7. But for experiment 6 it does not seem to be significant suggesting that experiment 6 lineups were difficult compared to the other experimental lineups. This shows a feature of the experimental designs where a mixture of difficult and easy lineups were included within the experiments as well as between the experiments.  For easy lineups there may be small chance to improve performances. But for harder lineups, the scope to improve performances is large.  Since for none of these experiments improvement in performances is observed, it is important in the sense that performances did not improve over attempts in both difficult and easy lineup situations.

@@

To visualize further how the performance in correct response improves over successive attempt, we fitted Model \eqref{eqn:trend_response} excluding the covariates related to attempt from the model and computed the residuals. The averages of these residuals for each of the attempts are plotted against attempt as shown in Figure \ref{fig:learning_trend_response}. Least square linear regression lines are drown through the points. Even though positive slopes over the attempts are observed, none of the trends is statistically significant. Subject specific least square lines shows large overall variability. Some subjects show improvement over time and some show the decrease in performance. The subject specific slopes are very different as well.

Model \eqref{eqn:trend_time} was compared to couple of other alternatives. A bigger model with attempt as a factor was considered but they did not appear to be significantly different with $p$ value XXX. A reduced model with attempt as a continuos covariate was also considered but they were significantly different with $p$-value XXX. Table \ref{tbl:model_result_time} presents the results of Model \eqref{eqn:trend_time}. The parameter $\alpha$ for fixed effect covariate attempt is highly significant in all the experiments. The negative estimates suggest that later attempt took less time for an evaluation. This suggest that even though observers did not improve the performance over attempt, they became efficient in responding faster. The parameter $\alpha_1$ for first attempt is also highly significant. The positive estimates of $\alpha_1$ indicates that first attempt made by an observer required much more times than other attempts. It is because for initial attempt the observer might have gone through instructions and became familiar with the experimental environment. Each page in the web site contains information about choice reason and the observers's confidence level. Also, the first page asks for observer Identification to be typed. The later pages of the web site was similar just the lineup was changed. Thus in the later attempts an observer does not need to spend any time for reading instructions. The model reflects that fact.

@@

\subsection{How People Pick the Data Plot} In each of the experiments we had a question for the turk observers to choose the reason for their selection of a particular data plot. It is shown by \cite{majumder:2013} that these choice reasons reveal the way people picks the data plot in the lineup. To investigate this further we added free text input option for their choice reason instead of some fixed reasons to select from. In this experiment people could write whatever they think their reasons for choice are. Figure \ref{fig:wordle} shows the words used to explain their reasons for choice. The most common words used to explain their choice are points and green which indicates two important feature of a plot. One is the indicator of plot aesthetics and the other is the color. Spread, steepest, line and apart are some other important words used frequently. Spread and apart are indicative of variability in the data. Steepest and line indicates some sort of systematic pattern in the data.
